---
title: "Task 2: Data Perspective on the evolution of education for 4- to 5-year-old
  children"
author: "P3 candidate"
date: "August 2024"
output:
  html_document: default
  pdf_document: default
---

<!-- set page width -->

```{=html}
<style type="text/css">
.main-container {
  max-width: 80% !important;
  margin: auto;
}
</style>
```
```{r setup, include=FALSE}
options(width=100)
knitr::opts_chunk$set(
	echo = FALSE,
	fig.height = 7,
	fig.width = 11,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)

library(tidyverse)
library(openxlsx)
library(writexl)
library(readxl)
library(janitor)
library(ggplot2)
library(plotly)
library(broom)
library(knitr)

# set wd
USERNAME    <- Sys.getenv("USERNAME")
USERPROFILE <- Sys.getenv("USERPROFILE")
wd          <- str_glue("{USERPROFILE}/Documents/Github/p3-assessment")

## input data ----
zwe <- read.csv(str_glue("{wd}/01_rawdata/Zimbabwe_children_under5_interview.csv"))
## input data ----

# the dataset includes variables used to produce the domains of the early childhood development index (ecdi), developed by UNICEF :: https://data.unicef.org/resources/early-childhood-development-index-2030-ecdi2030/

## funcs ----
# function to calculate indicator value
calculate_mean <- function(df) {
  df %>%
    mutate(lit_num_mean = mean(lit_num, na.rm = TRUE) * 100,
           phys_mean = mean(phys, na.rm = TRUE) * 100,
           learning_mean = mean(learning, na.rm = TRUE) * 100,
           soc_emo_mean = mean(soc_emo, na.rm = TRUE) * 100,
           ecdi_mean = mean(ecdi, na.rm = TRUE) * 100)
}
## funcs ----

## recode vars ----
df <- zwe %>%
  # positive indics
  mutate(across(c(EC6, EC7, EC8, EC9, EC11, EC12, EC13), ~ case_when(
    . == 1 ~ 1,
    . == 2 ~ 0,
    TRUE ~ NA_real_  
  )),
  # negative indics
  across(c(EC10, EC14, EC15), ~ case_when(
    . == 2 ~ 1,
    . == 1 ~ 0,
    TRUE ~ NA_real_  
  )))
## recode vars ----

## age in months ----
# support from chatgpt
df <- df %>%
  mutate(child_age_months = interval(child_birthday, interview_date) %/% months(1)) %>%
  relocate(child_age_months, .after = child_age_years)
## age in months ----

## educational areas ----
df2 <- df %>%
  mutate(lit_num = rowSums(select(., EC6, EC7, EC8), na.rm = TRUE),
         lit_num = ifelse(lit_num >= 2, 1, 0)) %>%
  # physical
  mutate(phys = ifelse((EC9 == 1 | EC10 == 1) & (!is.na(EC9) & !is.na(EC10)), 1, 0)) %>%
  # learning
  mutate(learning = rowSums(select(., EC11, EC12), na.rm = TRUE),
         learning = ifelse(learning >= 1, 1, 0)) %>%
  # socio-emotional
  mutate(soc_emo = rowSums(select(., EC13, EC14, EC15), na.rm = TRUE),
         soc_emo = ifelse(soc_emo >= 2, 1, 0))
## educational areas ----

## ecdi ----
# percentage of children who are developmentally on track in at least three of the four domains
df3 <- df2 %>%
  select(interview_date:child_birthday,
         lit_num,
         phys,
         learning,
         soc_emo) %>%
  mutate(ecdi = lit_num + phys + learning + soc_emo,
         ecdi = ifelse(ecdi >= 3, 1, 0))
## ecdi ----

## data (total) ----
# data for children 3-4 years (total)
av_total <- calculate_mean(df3) 

av_total <- av_total %>%
  mutate(child_age_years = "3-4 years") %>%
  select(child_age_years, contains("mean")) %>%
  distinct() %>%
  pivot_longer(-c(child_age_years), names_to = "indic", values_to = "value") %>%
  mutate(value_r = round(value, 0),
         indic = gsub("_mean", "", indic))
# compare with mics snapshot to ensure calculations are done correctly
## data (total) ----

## data (by month) ----
av_by_month <- df3 %>%
  group_by(child_age_months) %>%
  calculate_mean(.) %>%
  ungroup() 

av_by_month <- av_by_month %>%
  select(child_age_months, contains("mean")) %>%
  distinct() %>%
  pivot_longer(-c(child_age_months), names_to = "indic", values_to = "value") %>%
  mutate(value_r = round(value, 0),
         indic = gsub("_mean", "", indic)) %>%
  arrange(child_age_months, indic) %>%
  filter(!is.na(child_age_months))
## data (by month) ----



```

::: {style="text-align: center;"}
#### [[***Early Childhood Development Index (ECDI)***]{.underline}]{style="color: blue;"}
:::

# {.tabset}

## ECDI 

### Early Childhood Development Index (ECDI)

```{r plt1}
# prep data
plt1_dta <- av_by_month %>%
  mutate(country= "Zimbabwe") %>%
  filter(indic == "ecdi")

# plot
plt1 <- plt1_dta %>%
  ggplot(aes(x = child_age_months, y = value_r)) +
  geom_bar(data = plt1_dta, stat = "identity", fill = "#1CABE2",
            aes(text = str_c(country, 
                         '<br>Childs age (months): ', child_age_months,
                             '<br>ECDI (%): ', value_r, '<br>Year: 2019'
                             ))) +
  geom_smooth(method = "lm", se = FALSE, colour = "#002759") +
  labs(title = "Childhood development, by age in months, Zimbabwe, 2019", 
       x = "Age (months)", 
       y = "Percent (%)",
       caption = "Source: Re-analysis of Zimbabwe Multiple Indicator Cluster Survey 2019 microdata") +
  theme_minimal() + 
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 10)) +
  scale_x_continuous(limits = c(min(av_by_month$child_age_months)-1, max(av_by_month$child_age_months)+1), breaks = seq(min(av_by_month$child_age_months), max(av_by_month$child_age_months), by = 1)) +
  geom_text(data = plt1_dta %>% filter(child_age_months %in% c(36, 42, 48, 54, 59)),
    aes(label = value_r, y = value + 3),
    vjust = 0,
    size = 3.5,
    colour = "#1CABE2")

# make interactive
ggplotly(plt1, tooltip = "text")

# support from chatgpt:
# fit the linear model and extract summary
model <- lm(value_r ~ child_age_months, data = plt1_dta)
model_summary <- summary(model)
# coefficients table from the model summary
coefficients_summary <- as.data.frame(model_summary$coefficients)
# print in rmd
knitr::kable(coefficients_summary, caption = "Model Coefficients Summary")
```

-  This chart shows the early childhood development index (ECDI) by child’s age in months (blue bars), along with the line of best fit. The summary statistics are presented in the table below. 
-  The gradient of the line of best fit is positive, which indicates a positive relationship whereby as a child’s age increases, ECDI increases. Specifically, for each additional month a child has, ECDI is expected to increase by approximately `r round(coefficients_summary["child_age_months", "Estimate"], 2)` units.
- The p-value for this gradient is less than 0.05, indicating a statistically significant relationship and that the observed results are unlikely due to random chance. 
- This is important to know because if a child is getting older, but their ECDI is not increasing, there should be some intervention put in place to ensure the child is learning and thriving in their development.


## Domains 

### Educational domains that contribute to ECDI

```{r plt2}
plt2_dta <- av_by_month %>%
  filter(indic != "ecdi") %>%
  mutate(indic = case_when(indic == "learning" ~ "Learning",
                           indic == "lit_num" ~ "Literacy and math",
                           indic == "phys" ~ "Physical",
                           indic == "soc_emo" ~ "Socio-emotional"),
         country = "Zimbabwe")
  


plt2 <- ggplot(plt2_dta, aes(x = child_age_months, y = value_r, colour = indic)) +
  geom_point(aes(text = str_c(country, 
                         '<br>Childs age (months): ', child_age_months,
                         '<br>Domain: ', indic,
                         '<br>Percentage (%): ', value_r, 
                         '<br>Year: 2019'
                             ))) +
  geom_smooth(method = "lm", se = FALSE, colour = "black", size = 0.5) +
  labs(title = "Childhood development, by domain and age in months, Zimbabwe, 2019", 
       x = "Age (months)", 
       y = "Percent (%)",
       caption = "Source: Re-analysis of Zimbabwe Multiple Indicator Cluster Survey 2019 microdata") +
  theme_bw() +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20)) +
  scale_x_continuous(limits = c(min(av_by_month$child_age_months)-1, max(av_by_month$child_age_months)+1), breaks = seq(min(av_by_month$child_age_months), max(av_by_month$child_age_months), by = 1)) +
  facet_wrap(. ~ indic) +
  theme(legend.position="none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

# make interactive
ggplotly(plt2, tooltip = "text")

# support from chatgpt:
# fit a linear model for each group and extract summaries
model_summaries <- plt2_dta %>%
  group_by(indic) %>%
  do(tidy(lm(value_r ~ child_age_months, data = .))) %>%
  ungroup()

# reshape the data
model_summaries <- model_summaries %>%
  select(indic, term, estimate, std.error, statistic, p.value) %>%
  pivot_wider(names_from = term, values_from = c(estimate, std.error, statistic, p.value))

# print results
knitr::kable(model_summaries, caption = "Model Coefficients Summary by Group")

```

-  This chart shows the four educational domains that contribute to the ECDI, by child’s age in months, along with the line of best fit. The summary statistics are presented in the table below. 
-  The literacy and math domain had the strongest positive relationship with literacy and math development increasing 0.75 units per month increase in childs. This was statistically signficant (p<0.05), indicating that as a child's age increases, literacy and math skills increase the most of the four domains.
- Learning and socio-emotional domains also had significant positive relationships with child age in months. 
-  There was no statistically significant association found between childs age in months and physical activity (p=0.5).
-  Despite these observations, physical development was considerably higher throughout 36-59 months than literacy and math indicating that children generally develop physical skills at a younger age.
-  It is important to note that there are some variations in development indicator percentage, likely due to small sample sizes when breaking the average down by month of age. As such, results may be subject to noise and should be interpreted with this in mind.



